A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-09-01 demystify.tex

You can consider applying to MS programs, which are typically easier to get in; but you likely need to pay.  Sometimes, the university that rejected you for PhD might accept you for MS.
If you get into an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD. You will likely need to apply again, but you will have a much better chance because you have the direct support of a professor there.

@@

\section{What's the difference between PhD and MS in CS?}\label{sec:phd-vs-ms}
\sectioninfo{MS focuses on coursework and prepares you for industry, while PhD focuses on research and prepares you for academia or research labs.}

While both are graduate degrees, they are different in terms of objective, admission requirements, course requirements, duration, and funding as summarized in \autoref{tab:phd-vs-ms}.

  \item Objective: an MS is typically to prepare you for industry, while a PhD is to prepare you for research. Some MS has thesis option but the research is much lighter compared to a PhD, which is research-focused.

  \item Admission requirements: MS also requires a good GPA, LoRs, SOP, and test scores, but research experience is not required.  PhD requires all of these, but research experience is a must. Moreover, MS programs are typically much less competitive than PhD programs. In fact, many MS program are viewed as a revenue source (\emph{cash cow}) for the department, as they are not funded and students pay tuition.

  \item Funding: MS is typically \emph{not funded}, while PhD is typically funded (e.g., through TA or RA as explained in \autoref{sec:funding}).

@@

>>>>

You can consider applying to MS programs, which are typically easier to get in; but you likely need to pay (\autoref{sec:ms-admission}).  A university that rejected you for PhD might accept you for MS.
If you get into an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD. You will likely need to apply again, but you will have a much better chance because you have the direct support of a professor there.

@@

@@

\chapter{MS Admission}
\chapterinfo{MS focuses on coursework and prepares you for industry, while PhD focuses on research and prepares you for academia or research labs. This section provides guidance on applying to MS programs.}

While both MS and PhD programs are graduate degrees, they are \emph{very} different in terms of objective, admission requirements, course requirements, duration, and funding. This section provides guidance on applying to MS programs in CS.

\section{Differences between PhD and MS}

\autoref{tab:phd-vs-ms} summarizes the main differences between PhD and MS programs:
\begin{itemize}

  \item Objective: an MS is typically to prepare you for industry, while a PhD is to prepare you for research. Some MS has thesis option but the research is much lighter compared to a PhD.

  \item Admission requirements (\autoref{sec:ms-admission}): MS also requires a good GPA, LoRs, SOP, and test scores, but research experience is not required.  PhD requires all of these, but research experience is a must.

  \item Funding: MS is typically \emph{not funded}, while PhD is typically funded (\autoref{sec:funding}).

\section{MS Admission}\label{sec:ms-admission}

Except for a very few highly selective programs, MS programs are typically much less competitive than PhD programs. Some refer to MS programs as a \emph{cash cow} for the department, as MS students are often not funded and have to pay tuition (i.e., you're likely to get in if you can afford it).

While admission requirements are similar to PhD programs (e.g., GPA, SOP, LORs), research is not a focus in MS programs. In most cases, the main requirements are that you have sufficient background in CS (e.g., through your undergrad degree).

\subsection{Admission Committee}
MS admission also involves an adcom that reviews applications and makes admission decisions. However, MS admission is typically rolling (i.e., they review applications as they come in and make decisions throughout the admission cycle).  Moreover, unlike PhD that has multiple reviewers for each application, an MS application typically involves only one reviewer and does not review much time to review compared to a PhD application (\autoref{sec:ievaluate}).

\begin{commentbox}[Vu:]
    Mason has a very large the number of MS students in CS (1000+ MS vs 200 PhDs). In contrast, other similar size universities often have much smaller MS CS programs (or none). Location plays an important role as GMU is close to DC with many developer professionals who want to get an MS, which are often covered by their employers.  Our MS program is geared towards working professionals, e.g., all of our MS courses are offered in the evening and online.

    The CS department has three separate committees for MS admission: MS in CS (the traditional one, which is the largest), MS in Software Engineering (SWE), and MS in Information System. 
    Each committee has its own chair and members. For example, I am often in the MS SWE adcom, which has about 3 adcom members (including the chair who also does the review). 
\end{commentbox}

%Third, adcom members can involve both teaching and tenure-line faculty (\autoref{sec:faculty-types}).

\subsection{Application Materials}

In an MS application, you typically need to submit the similar materials as in a PhD application, e.g., transcripts, LoRs, SOP, and test scores.  However, research experience is not required, and LoRs can be from anyone who can speak about your academic or working abilities (e.g., professors, employers, etc). Moreover, your SOP you should focus on your academic and work (e.g., software development) background, why you want to get an MS, and how the program fits your goals. As with PhDs, GREs are often not required.  Moreover, some MS programs do not even require LoRs or SOPs.

\paragraph{Transcripts and Undergrad Background} Since research is not required, your undergrad background is more important in an MS application.  You should have a strong background in CS, e.g., through your undergrad degree.  If you do not have a CS degree, you might need to take some prerequisite courses before you can start the MS program.  GPA is also important, as it is often used as a filter for MS applications.

\paragraph{SOP} Your SOP should explain why you want to get an MS, how the program fits your goals, and why you're a good fit for the program.  You don't need to mention about working with specific professors unless you want to do a thesis option. You still need to customize it for each school (e.g., you picked GMU due to its strong SWE program or the DC area has many job opportunities).

\paragraph{LoRs} LoRs can be from professors, employers, or anyone who can speak about your academic or working abilities.  Unlike PhDs, your LoRs do not need to be from professors and talk about research experience. Many schools do not even require LoRs for MS.

\paragraph{Test Scores} GREs are often not required for MS applications.  However, some schools might still require them, and you should check with the school you're applying to.  If you have a low GPA, a high GRE score can help offset that.

### 2024-09-01 faculty.tex

\chapter{Teaching}

While teaching should take about 40\% of your time (equivalent to research) and your salary is mainly for teaching, teaching is not as nearly as important as research for tenure.  Moreover, for most new faculty, teaching is not as difficult as getting grants or publishing papers. Thus, tenure-track faculty often do not worry much about teaching.

\section{Teaching Load}
\section{Teaching Evaluation}

>>>>

\chapter{Teaching}\label{chapter}

While teaching should take about 40\% of your time (equivalent to research) and your salary is mainly for teaching, it is not as nearly as important as research for tenure.  Moreover, for most new faculty, teaching is not as difficult as getting grants or publishing papers, and thus is often taken lightly by new faculty.  However, teaching is still important for your tenure case, and being a good teacher can be very rewarding. Moreover, innovative integration of research and teaching can help you win good students and even NSF grants.

\section{Teaching Load}
New TT faculty often have a lighter teaching load in their first several years or even throughout their tenure track.  This is to give you time to establish your research program.  However, you should expect to teach at least one course per semester.  The teaching load is usually higher at teaching universities, and lower at research universities. At some universities, you might be able to buy out of teaching by getting a grant.

\section{Teaching Evaluation}
